---
title: "Complex Spatial Priors"
author: "Michael Stevens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{"Complex Spatial Priors"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
library(silverblaze)
library(raster)
library(rgdal)
plasma <- c("#F0F921FF", "#FDC926FF", "#FA9E3BFF", "#ED7953FF", "#D8576BFF",
            "#BD3786FF", "#9C179EFF", "#7301A8FF", "#47039FFF", "#0D0887FF")
set.seed(2)
```

## Introduction
The geographic profiling models described in the previous tutorials all make use of a Metropolis-Hastings algorithm for updating source locations. As a result, we can impose any prior we like on source locations. So far, priors on source locations have assumed that the probability of observing a source at a specific location is either constant (inside the shapefile) or zero (outside the shapefile). When we update source locations via the Metropolis-Hastings algorithm any proposed value that is outside the shapefile is immediately rejected. This differs from the conventional geographic profiling model (see @Verity2014 and @Faulkner2017) that implements the shapefile post-hoc. This tutorial will walk through different options for spatial priors in the silverblaze package. Let's start with a refresher: a uniform prior based on simulated data. The prior on source locations is defined over a grid of cells, hence we create a uniform prior via the `raster_grid()` function.    

```{r}
# define sentinel site array
sentinal_lon <- seq(-0.2, 0.0, l = 15)
sentinal_lat <- seq(51.45, 51.55, l = 15)
sentinal_grid <- expand.grid(sentinal_lon, sentinal_lat)
names(sentinal_grid) <- c("longitude", "latitude")

# create example dataset
data <- sim_data(sentinel_lon = sentinal_grid[,1],
                 sentinel_lat = sentinal_grid[,2],
                 sentinel_radius = 0.3, 
                 K = 2,
                 source_lon_min = min(sentinal_lon), 
                 source_lon_max = max(sentinal_lon), 
                 source_lat_min = min(sentinal_lat),
                 source_lat_max = max(sentinal_lat), 
                 sigma_model = "single", 
                 sigma_mean = 1, 
                 sigma_var = 0,
                 expected_popsize = 1e3)
                 
# store objects for easy access later
df_all <- data$data
dataPoints <- data$record$data_all
true_sources <- data$record$true_source
```

Now let's create a raster based on the data we've just simulated, this will work as a foundation for our spatial prior.  

```{r}
# create uniform prior on source locations
raster_frame <- raster_grid(cells_lon = 100, cells_lat = 100,
                            range_lat = range(sentinal_lat),
                            range_lon = range(sentinal_lon), guard_rail = 0.25)
```

This spatial prior is then fed into our geographic profiling project using the `spatial_prior` argument in the `new_set()` function. We then specify which kind of prior we wish to impose on source locations, using this raster, via the `source_model` argument.  

```{r}
# create GP project
p <- rgeoprofile_project()
p <- bind_data(p,
               df = df_all,
               data_type = "counts")
  
# add parameter set to project
p <- new_set(project = p,
             spatial_prior = raster_frame,
             source_model = "uniform",
             sentinel_radius = 0.3,
             sigma_model = "single",
             sigma_prior_mean = 2,
             sigma_prior_sd = 1,
             expected_popsize_prior_mean = 1e3,
             expected_popsize_prior_sd = 1e2, 
             name = "uniform_prior_project")
```

Now let's have a look at our data and spatial prior. 

```{r, echo = FALSE}

plot2 <- plot_map(map_type = 110)
plot2 <- overlay_spatial_prior(plot2, p, opacity = 0.5)
plot2 <- overlay_sentinels(plot2, 
                           p, 
                           border = F,
                           legend = FALSE, 
                           fill_opacity = 1, 
                           sentinel_radius = 0.3)
plot2 <- overlay_sources(plot2, 
                         true_sources$longitude, 
                         true_sources$latitude, 
                         icon_width = 15, 
                         icon_height = 15)

plot2

```

## Bi-variate normal prior

A more informative prior could be a bivariate normal distribution centred on the spatial mean of the positive data. Similarly to @Verity2014, the standard deviation of this bivariate normal prior is set to the maximum distance bewteen the mean and the positive data.  

```{r}
# add parameter set to project
p <- new_set(project = p,
             spatial_prior = raster_frame,
             source_model = "normal",
             sentinel_radius = 0.3,
             sigma_model = "single",
             sigma_prior_mean = 2,
             sigma_prior_sd = 1,
             expected_popsize_prior_mean = 1e3,
             expected_popsize_prior_sd = 1e2, 
             name = "bvnorm_prior_project")
```

```{r, echo = FALSE}

plot2 <- plot_map(map_type = 110)
plot2 <- overlay_spatial_prior(plot2, p, opacity = 0.5)
plot2 <- overlay_sentinels(plot2, 
                           p, 
                           border = F,
                           legend = FALSE, 
                           fill_opacity = 1, 
                           sentinel_radius = 0.3)
plot2 <- overlay_sources(plot2, 
                         true_sources$longitude, 
                         true_sources$latitude, 
                         icon_width = 15, 
                         icon_height = 15)

plot2

```

## Kernel density prior

A bivariate normal distribution may be an inappropriate choice for a prior if the spatial mean of the positive data is an area containing absences. Hence, another option for `source_model` is `"kernel".` This will produce a kernel density estimate for the positive data (see @Barnard2010 for a description of the KDE and bandwidth choice). 

```{r}
# add parameter set to project
p <- new_set(project = p,
             spatial_prior = raster_frame,
             source_model = "kernel",
             sentinel_radius = 0.3,
             sigma_model = "single",
             sigma_prior_mean = 2,
             sigma_prior_sd = 1,
             expected_popsize_prior_mean = 1e3,
             expected_popsize_prior_sd = 1e2, 
             name = "kde_prior_project")
```

```{r, echo = FALSE}

plot2 <- plot_map(map_type = 110)
plot2 <- overlay_spatial_prior(plot2, p, opacity = 0.5)
plot2 <- overlay_sentinels(plot2, 
                           p, 
                           border = F,
                           legend = FALSE, 
                           fill_opacity = 1, 
                           sentinel_radius = 0.3)
plot2 <- overlay_sources(plot2, 
                         true_sources$longitude, 
                         true_sources$latitude, 
                         icon_width = 15, 
                         icon_height = 15)

plot2

```

## Using shapefiles
We built our raster from scratch using the `raster_grid()` function, however, silverblaze also allows for the importing of shapefiles such that the boundaries of our raster conform to a particular geographic boundary. The example here relates to a particular borough of London.

```{r, message = FALSE, warning = FALSE, error = FALSE}
s <- rgeoprofile_shapefile("London_north")
spatial_prior <- raster_from_shapefile(s, cells_lon = 100, cells_lat = 100)
plot(spatial_prior, xlab = "Longitude", ylab = "Latitude")
```

This shapefile is built into the package, any other shapefile should be loaded using the `readOGR()` function from [rgdal](https://cran.r-project.org/web/packages/rgdal/index.html).   

## A non-uniform spatial prior

There are a lot more shapefiles out there that will provide more information that we might want to influence our model. For example the [London Datastore](https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london) provides shapefiles with various spatial information. 

```{r}
dataStore <- rgeoprofile_shapefile("London_boundary/ESRI")
print(dataStore)
```

Here we see a summary of the data. Lets extract population density from this shapefile and convert it into a raster to use as our spatial prior.

## Converting a shapefile to a raster

```{r}
# Firstly we create a raster that matches the extent of the shapefile.
newRaster <- raster(ncol = 100, nrow = 100)
extent(newRaster) <- extent(dataStore)

# We then rasterize our shapefile using the raster we just created.
PopDenRaster <- rasterize(dataStore, newRaster, field = "POPDEN")

# Finally we project the raster to the correct co-ordinate reference system (CRS).
crs(PopDenRaster) <- crs(dataStore)
PopDenRaster <-  projectRaster(PopDenRaster, crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
summary(values(PopDenRaster))

# We've obtained our raster, now we need to normalise the population density values so that we're working with probabilities.
values(PopDenRaster) <- values(PopDenRaster)/sum(values(PopDenRaster)[!is.na(values(PopDenRaster))])

# Let's take a look at what the final results. 
plot(PopDenRaster, xlab = "Longitude", ylab = "Latitude")
```

Now we've loaded the shapefile and converted it into a spatial prior, let generate some data, and run the model using the new prior.  

## A non-uniform spatial prior in a GP project

```{r}
set.seed(10)
# create example dataset
Extent <- extent(PopDenRaster)
sentGrid <- expand.grid(seq(Extent[1] + 0.1, Extent[2] - 0.1, l = 10), seq(Extent[3]+ 0.05, Extent[4]- 0.05, l = 15))
data <- sim_data(sentinel_lon = sentGrid[,1],
                 sentinel_lat = sentGrid[,2],
                 sentinel_radius = 0.3, 
                 K = 2,
                 source_lon_min = Extent[1] + 0.1, 
                 source_lon_max = Extent[2] - 0.1, 
                 source_lat_min = Extent[3] + 0.05,
                 source_lat_max = Extent[4] - 0.05, 
                 sigma_model = "single", 
                 sigma_mean = 4, 
                 sigma_var = 0,
                 expected_popsize = 1000)
df_all <- data$data
dataPoints <- data$record$data_all
true_sources <- data$record$true_source
```

Create a geographic profiling project:

```{r}
pPopDen <- rgeoprofile_project()
pPopDen <- bind_data(pPopDen,
                     df = df_all,
                     data_type = "counts")
```

Create parameter sets:

```{r}
# add parameter set to project
pPopDen <- new_set(project = pPopDen,
                   spatial_prior = PopDenRaster,
                   source_model = "manual",
                   sentinel_radius = 0.3,
                   sigma_model = "single",
                   sigma_prior_mean = 4,
                   sigma_prior_sd = 0.1,
                   expected_popsize_prior_sd = 1e3)
```

Plot the data with the prior:

```{r, echo = FALSE}
plot1 <- plot_map(map_type = 110)
plot1 <- overlay_sentinels(plot1, pPopDen, sentinel_radius = 1)
plot1 <- overlay_sources(plot1, lon = true_sources$longitude, lat = true_sources$latitude)
plot1 <- overlay_points(plot1, lon = dataPoints$longitude, lat = dataPoints$latitude)
plot1 <- overlay_spatial_prior(plot1, pPopDen, opacity = 0.5)
plot1
```

Run the MCMC: 

```{r, eval = FALSE}
set.seed(10)
pPopDen <- run_mcmc(project = pPopDen,
                    K = 1:2,
                    burnin = 2e4,
                    samples = 2e4,
                    pb_markdown = TRUE,
                    converge_test = 1e4)
```

```{r, echo = FALSE}
pPopDen <- rgeoprofile_file("complex_spatial_project.rds")
```

Plot the prior and the geoprofile:

```{r}
# uniform
plot3 <- plot_map()
plot3 <- overlay_sources(plot3, 
                         lon = true_sources$longitude, 
                         lat = true_sources$latitude)
                         
plot3 <- overlay_geoprofile(plot3, 
                            pPopDen, 
                            K = 2, 
                            smoothing = 2, 
                            threshold = 1,
                            col = plasma,
                            legend = TRUE)
                            
plot_loglike_diagnostic(pPopDen, K = 2)
plot3
```

```{r, echo = FALSE, eval = FALSE}
# secretly save/load data from file so output is consistent between tutorials
# saveRDS(pPopDen, file = "/home/mstevens/Desktop/MAIN WORK/Presence Absence/Package Versions/silverblaze/inst/extdata/complex_spatial_project.rds")
```
## References
