---
title: "Tutorial 3: Model Validation"
author: "Michael Stevens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: model-validation.bib
vignette: >
  %\VignetteIndexEntry{"Model Validation"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(silverblaze)
p <- rgeoprofile_file("tutorial1_project.rds")
```
The majority of the following methodology on Bayesian model validation is adopted from @Spiegelhalter2014.
There are many functions silverblaze has to offer for checking the ability of the model but we've done very little to compare the difference between models fitting different values of `K`. Hence we introduce the `plot_DIC_gelman()` function to calculate the deviance information criterion (DIC) for each model. Let's have a look at the DIC values for our models from [tutorial 1](https://michael-stevens-27.github.io/silverblaze/articles/basic-implementation.html).

```{r}
plot_DIC_gelman(p)
```

As we can see the model with the lowest DIC is the one fitting 3 sources. This is exactly what we would expect given we simulated under a model where `K = 3`. 

### References
