---
title: "Tutorial 3: Model Validation"
author: "Michael Stevens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: model-validation.bib
vignette: >
  %\VignetteIndexEntry{"Tutorial 3: Model Validation"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(silverblaze)
p <- rgeoprofile_file("tutorial1_project.rds")
```
The majority of the following methodology on Bayesian model validation is adopted from @Spiegelhalter2014.
There are many functions silverblaze has to offer for checking the ability of the model but we've done very little to compare the difference between models fitting different values of `K`. Hence we introduce the `plot_DIC_gelman()` function to calculate the deviance information criterion (DIC) for each model.

```{r}
plot_DIC_gelman(p)
```

As we can see the model with the lowest absolute DIC is the one fitting XXXX sources. This of course should not be the case indicating the DIC is an unreliable metric for this analysis. 

### References
