---
title: "Variable Expected Population"
author: "Michael Stevens"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Variable-sigma.bib
vignette: >
  %\VignetteIndexEntry{"Variable Expected Population"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

A variable sigma model is available within the Silverblaze package but is yet to be investigated. The presence-absence manuscript focusses on the development of the model and assumes what all literature around GP has thus far assumed: the dispersal associated with an offender, an invasive species, or infectious disease is the same across source locations. @Ratcliffe2006 makes it clear that this is not always the case. An offender that commits crime around a work place will be restricted by time contstraints to return to work within a certain time. This compares to more relaxed time contraints when commiting crimes around a home. So we might expect a tight dispersal around a workplace but a wider dispersal around a home. Here we generate a couple of toy examples to investigate the following questions:

* Can the model successfully return different dispersal (sigma) values for different source locations?
* How well does the model perform when there is overlap in the dispersal ranges of sources with different sources?        

The first question will illustrate how one sets up the variable (independent) sigma model in the case where we have one large sigma and one small sigma across two sources. The second question looks to investigate any mixing issues when the dispersal ranges across sources overlap.

Load Silverblaze package:

```{r, message=FALSE}
# load packages
library(silverblaze)
# set seed
set.seed(1)
```

# Simple two source example

## Setting up and running the indpendent sigma model

Set up an array of sentinel sites and a uniform prior. We'll also specify the sentinel site radius. 

```{r}

sentinal_lon <- seq(-0.2, 0.0, l = 12)
sentinal_lat <- seq(51.45, 51.55, l = 12)
sentinal_grid <- expand.grid(sentinal_lon, sentinal_lat)
names(sentinal_grid) <- c("longitude", "latitude")

uniform_prior <- raster_grid(cells_lon = 100, cells_lat = 100,
                               range_lat = range(sentinal_lat),
                               range_lon = range(sentinal_lon), guard_rail = 0.5)
                               
sentinel_site_radius <- 0.2                               
```

Now we generate some data. We use the `sim_data()` function to generate two sources, speicifying the kind of sigma model we want to draw from by setting `sigma_model = "indpendent"`. For each source we choose a sigma value using `sigma_mean` and `sigma_var`. This specifies the mean and variance of each normal distribution a sigma is being drawn from.   

```{r}
sim1 <- sim_data(sentinal_grid$longitude,
                 sentinal_grid$latitude,
                 sigma_model = "independent", # set the sigma model to indpendent
                 sigma_mean = c(1, 1.5),      
                 sigma_var = c(0, 0),         
                 sentinel_radius = sentinel_site_radius,
                 K = 2,
                 expected_popsize = 1000)

data_all1 <- sim1$record$data_all
true_source1 <- sim1$record$true_source
K_model <- 2

# pull the different sigma values
true_sigma1 <- sim1$record$true_sigma
```

Notice `sigma_var` is equal to zero for both sources, this allows us to force the value of each sigma. Now we build our geoprofiling project, add in our data and setup parameters to run the MCMC. 

```{r}

# create a project and bind data
p1 <- rgeoprofile_project()
p1 <- bind_data(p1,
                df = sim1$data,
                data_type = "counts")

# add parameter set to project
p1 <- new_set(project = p1,
             spatial_prior = uniform_prior,
             sentinel_radius = sentinel_site_radius,
             sigma_model = "independent",
             sigma_prior_mean = 1,
             sigma_prior_sd = 100,
             expected_popsize_prior_mean = 1000,
             expected_popsize_prior_sd = 500)

plot1 <- plot_map()
plot1 <- overlay_points(plot1, data_all1$lon, data_all1$lat)
plot1 <- overlay_sources(plot1, true_source1$lon, true_source1$lat)
plot1
```

We can see a source with a wider sigma and if we zoom into the bottom left we can see a source with auch tighter sigma. Let's create our geoprofiling project, load in this data and run the MCMC under a variable sigma model:

```{r, eval = FALSE}
# run MCMC under K=2 model
p1 <- run_mcmc(project = p1,
              K = K_model,
              burnin = 5e4,
              samples = 5e4,
              converge_test = 1e4)
```
```{r, echo = FALSE}
p1 <- rgeoprofile_file("VarSigproject_no_overlap.rds")
```
## Results
Below we can see the geoprofile of the two sources. Just by eye it looks like each cluster has a different sigma value.  

```{r}
# produce and store map
plot1 <- plot_map()
plot1 <- overlay_sentinels(plot1, p1, sentinel_radius = sentinel_site_radius)
plot1 <- overlay_geoprofile(plot1, p1, K = 2, smoothing = 2)
plot1 <- overlay_sources(plot1, true_source1$lon, true_source1$lat)
plot1
```

How can we be sure these sigma values are different? We can access the raw output associated to sigma using the `get_output()` function. We can also look into where our true values fall within the 95% confidence intervals of the fitted values.

```{r}
# get the raw sigma draws
sigma_output <- get_output(p1, "sigma_sampling", K = 2, type = "raw")
sigma1 <- density(sigma_output[,1])
sigma2 <- density(sigma_output[,2])

par(mfrow = c(1,2))
plot(sigma1, main = "Density of Sigma 1")
plot(sigma2, main = "Density of Sigma 2")

# calculate the mean of these draws
estimatedSigma <- apply(sigma_output, 2, mean)
estimatedSigma

# get the confidence intervals for sigma
sigma_intervals <-get_output(p1, "sigma_intervals", K = 2)
sigma_intervals

# print true values 
true_sigma1
```

# Overlapping sources and variable dispersal 

Now we generate another data set, both sources will be closer than the previous data to investigate the models ability to pick up two sources of varying dispersal on top of one another. 

```{r}
sim2 <- sim_data(sentinal_grid$longitude,
                 sentinal_grid$latitude,
                 source_lon_min = -0.1, 
                 source_lon_max = -0.1,
                 source_lat_min = 51.49,
                 source_lat_max = 51.51,                 
                 sigma_model = "independent", # set the sigma model to indpendent
                 sigma_mean = c(0.5, 2),
                 sigma_var = c(0.1, 0.1),
                 sentinel_radius = sentinel_site_radius,
                 K = 2,
                 expected_popsize = 1000)

data_all2 <- sim2$record$data_all
true_source2 <- sim2$record$true_source

# pull the different sigma values
true_sigma2 <- sim2$record$true_sigma

# create a project and bind data
p2 <- rgeoprofile_project()
p2 <- bind_data(p2,
                df = sim2$data,
                data_type = "counts")

# add parameter set to project
p2 <- new_set(project = p2,
            spatial_prior = uniform_prior,
            sentinel_radius = sentinel_site_radius,
            sigma_model = "independent",
            sigma_prior_mean = 1,
            sigma_prior_sd = 100,
            expected_popsize_prior_mean = 1000,
            expected_popsize_prior_sd = 500)       
```

Plot the the new data:

```{r}
plot2 <- plot_map()
plot2 <- overlay_points(plot2, data_all2$lon, data_all2$lat)
plot2 <- overlay_sources(plot2, true_source2$lon, true_source2$lat)
plot2
```

Run the MCMC and plot the new geoprofile

```{r, eval = FALSE}
# run MCMC under K = 2 model
p2 <- run_mcmc(project = p2,
              K = K_model,
              burnin = 1e4,
              samples = 1e4,
              converge_test = 1e4)
```
```{r, echo = FALSE}
p2 <- rgeoprofile_file("VarSigproject_overlap.rds")
```

Below we can see the geoprofile of the two sources.
  
```{r, echo = FALSE}
# produce and store map
plot2 <- plot_map()
plot2 <- overlay_sentinels(plot2, p2, sentinel_radius = sentinel_site_radius)
plot2 <- overlay_geoprofile(plot2, p2, K = 2, smoothing = 2)
plot2 <- overlay_sources(plot2, true_source2$lon, true_source2$lat)
plot2
```

## Results

Again let's check the raw output of the MCMC chain for each sigma. 

```{r}
# get the raw sigma draws
sigma_output2 <- get_output(p2, "sigma_sampling", K = 2, type = "raw")
sigma1 <- density(sigma_output2[,1])
sigma2 <- density(sigma_output2[,2])

plot(sigma1, main = "Density of Sigma 1")
plot(sigma2, main = "Density of Sigma 2")

# calculate the mean of these draws
estimatedSigma2 <- apply(sigma_output2, 2, mean)
estimatedSigma2

# get the confidence intervals for sigma
sigma_intervals2 <-get_output(p2, "sigma_intervals", K = 2)
sigma_intervals2

# print true values 
true_sigma2
```
### References

```{r, echo = FALSE, eval = FALSE}
# secretly save/load data from file so output is consistent between tutorials
#saveRDS(p1, file = "../inst/extdata/VarSigproject_no_overlap.rds")
#saveRDS(p2, file = "./inst/extdata/VarSigproject_overlap.rds")
```
